### 2022-06-20 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@a215c933433f5547874781355eed948badd49b0e 🚀

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs (Plowright et al. 2017). Although host richness is often used
as a superficial proxy for spillover risk (Anthony et al. 2017,
Ruiz-Aravena et al. 2022), these approaches oversimplify the relevant
interspecific heterogeneity in immunology, behavior, and other traits,
and therefore overlook unique host pools that allow for the rapid
evolution of highly divergent viruses (Agosta et al. 2010). In the case
of generalist pathogens like betacoronaviruses, there is conceptual and
empirical support to the idea that these community-level mechanisms are
even more important (Power and Mitchell 2004). This creates a disconnect
between coevolutionary theory and ecological studies of spillover risk.

The Geographic Mosaic Theory of Coevolution (GMTC; Thompson 2005)
attempts to explicitly connect microevolutionary dynamics to the
macroecology and biogeography of symbiotic interactions. The GMTC posits
that coevolutionary processes between pairs (Thompson 1994) or complexes
(Janzen 1980) of species are structured in space by the rippling effects
of abiotic conditions onto evolutionary mechanism, giving rise to
fragmented systems with different structure and ecologically dynamics
over large spatial extents (see \emph{e.g.} Price 2002). The GMTC
predicts a spatial fragmentation of coevolutionary dynamics under the
joint action of three processes (see notably Gomulkiewicz et al. 2007):
coevolutionary hot- and coldspots, which appear when the intensity of
\emph{interaction} (in terms of reciprocal fitness consequences) varies
spatially; selection mosaics, wherein the intensity of \emph{selection}
varies across space, driven by both the biotic complexity of the
community (locally diverse hosts and viruses are more biotically
complex) and the local favorability of the environment
(\textbf{Thrall2007?}); and trait remixing, which occurs when
coevolutionary dynamics are driven by by the arrival (or departure) of
functional traits, through changes in community composition due to
invasions, meta-community dynamics, and disperal.

Here, we develop a framework that quantifies each of these elements in
light of a measurable aspect of host-virus interactions. To do so, we
explore the global biogeography of betacoronaviruses, the group that
includes SARS-CoV, MERS-CoV, and SARS-CoV-2. Working from a recently
published database of bat hosts of betacoronaviruses, we develop the
first global maps of both host and virus evolutionary distinctiveness
and biogeographic regions for this system. Aiming to explain these
patterns, we develop a generalized framework for applying the GMTC to
host-virus interactions, with a specific emphasis on the potential to
create independent coevolutionary dynamics (and therefore spatial
fragmentation in risk) through heterogeneity. We develop a trivariate
risk assessment system that connects each GMTC mechanism o a
quantifiable aspect of host-virus interactions: (i) viral sharing rates
in host communities, representing the strength of potential interaction
between viruses and any one host (i.e., places where viruses undergo
constant host switching may be coevolutionary coldspots); (ii) the
phylogenetic diversity of hosts, as a proxy for variation in the
immunological mechanisms that antagonize viruses (i.e., the selection
mosaic); and (iii) the local uniqueness of the bat community,
representing the potential for viruses to be exposed to novel host
traits (e.g., variation in receptor sequences). Together, we argue that
these can be used to identify and map the evolutionary drivers that---in
conjunction with transmission processes (e.g., viral prevalence in
reservoirs and animal-human contact rates)--- determine disease
emergence risk.

>>>>

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs (Plowright et al. 2017). Although host richness is often used
as a superficial proxy for spillover risk (Anthony et al. 2017,
Ruiz-Aravena et al. 2022), these approaches oversimplify the relevant
interspecific heterogeneity in immunology, behavior, and other traits,
and therefore overlook unique host pools that allow for the rapid
evolution of highly divergent viruses (Agosta et al. 2010). In the case
of generalist pathogens like betacoronaviruses, there is conceptual and
empirical support to the idea that these community-level mechanisms are
even more important (Power and Mitchell 2004). This creates a disconnect
between coevolutionary theory and existing ecological frameworks for
mapping spillover risk.

The Geographic Mosaic Theory of Coevolution (GMTC; Thompson 2005)
attempts to explicitly connect microevolutionary dynamics to the
macroecology and biogeography of symbiotic interactions. The GMTC posits
that coevolutionary processes among pairs (Thompson 1994) or complexes
(Janzen 1980) of species are structured in space by the rippling effects
of abiotic conditions onto evolutionary mechanism, giving rise to
fragmented systems with different structure and ecologically dynamics
over large spatial extents (see \emph{e.g.} Price 2002). The GMTC
predicts a spatial fragmentation of coevolutionary dynamics under the
joint action of three processes (see notably Gomulkiewicz et al. 2007):
coevolutionary hot- and coldspots, which appear when the intensity of
\emph{interaction} (in terms of reciprocal fitness consequences) varies
spatially; selection mosaics, wherein the intensity of \emph{selection}
varies across space, driven by both the biotic complexity of the
community (locally diverse hosts and viruses are more biotically
complex) and the local favorability of the environment
(\textbf{Thrall2007?}); and trait remixing, which occurs when
coevolutionary dynamics are driven by by the arrival (or departure) of
\emph{functional traits}, through changes in community composition due
to invasions, meta-community dynamics, and disperal.

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. Working from a recently published database of bat hosts of
betacoronaviruses, we develop the first global maps of both host and
virus evolutionary distinctiveness and biogeographic regions for this
system. Aiming to explain these patterns, we develop a generalized
framework for applying the GMTC to host-virus interactions, with a
specific emphasis on the potential to create independent coevolutionary
dynamics (and therefore spatial fragmentation in risk) through
heterogeneity. We develop a trivariate risk assessment system that
connects each GMTC mechanism to a quantifiable aspect of host-virus
interactions: (i) viral sharing rates in host communities, representing
the strength of potential interaction between viruses and any one host
(i.e., places where viruses undergo constant host switching may be
coevolutionary coldspots); (ii) the phylogenetic diversity of hosts, as
a proxy for variation in the immunological mechanisms that antagonize
viruses (i.e., the selection mosaic); and (iii) the local uniqueness of
the bat community, representing the potential for viruses to be exposed
to novel host traits (e.g., variation in receptor sequences). Together,
we argue that these can be used to identify and map the evolutionary
drivers that---in conjunction with transmission processes (e.g., viral
prevalence in reservoirs and animal-human contact rates)--- determine
disease emergence risk.
